---
title: "Using spaceNtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spaceNtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package spaceNtime was created to assist with analyses described in ["Three novel methods to estimate abundance of unmarked animals using remote cameras" by Moeller et al. (2018)](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2331). This package helps the user create an encounter history from a database of analyzed photos and run each of the models.

Currently, this package only includes STE, but the other methods are on their way. 

This is my first package and is very much in development, so it is guaranteed to have bugs. I would appreciate hearing from you if you encounter problems!

## Load the package
If you haven't already installed it, please see instructions at the bottom of my [Github repository](https://github.com/annam21/spaceNtime)
```{r, message = F}
library(spaceNtime)
```


## The data

The functions for all three methods require two pieces of information.

### Photo information

The first, called df, is a data.frame or tibble that contains all your pictures from the study period. This must contain, at a minimum, these three columns (names and classes must match exactly): 

- cam: The unique ID of the sampling site (any class)

- datetime: The date and time of the photo (class POSIXct)

- count: The count of your study species at each date and time (class numeric)

```{r}
df <- data.frame(
  cam = c(1,1,2,2,2),
  datetime = as.POSIXct(c("2016-01-02 12:00:00",
                          "2016-01-03 13:12:00",
                          "2016-01-02 12:00:00",
                          "2016-01-02 14:00:00",
                          "2016-01-03 16:53:42"),
                        tz = "GMT"),
  count = c(1, 0, 0, 1, 2)
)
```
`r knitr::kable(df, caption = "df")`

### Camera deployment information

The user must also provide a database called deploy. This is a dataframe or tibble that contains information about the active periods for each active camera and their areas. **Please note,** to get accurate estimates, this should include all cameras that were functioning, not just the ones that got pictures of your target species. 

There are two possible versions of deploy. Studies with timelapse photos can use either version. 

#### Version 1: timelapse only
```{r, echo = F}
dep1 <- data.frame(
  cam = c(1,1,2,2,2),
  datetime = as.POSIXct(c("2016-01-02 12:00:00",
                          "2016-01-02 14:00:00",
                          "2016-01-02 12:00:00",
                          "2016-01-02 14:00:00",
                          "2016-01-03 16:00:00"),
                        tz = "GMT"),
  area = c(300, 300, 200, 0, 450)
)
```
`r knitr::kable(dep1, caption = "deploy version 1")`

This version of deploy will look similar or identical to df, with the column "area". If should contain *every* photo taken during the study period. It will have the following columns, at a minimum:

- cam: The unique ID of the sampling site (same class as df$cam)

- datetime: The date and time of the photo (same class and tz as df$datetime)

- area: the camera's visible area, in any units (class numeric)

The area should be 0 on any occasion that the camera was obstructed or malfunctioning. 
We will later look for any missing photos, which will indicate the camera was not functioning properly. 

#### Version 2: timelapse or motion-sensor

```{r}
deploy <- data.frame(
  cam = c(1, 2, 2, 2),
  start = as.POSIXct(c("2015-12-01 15:00:00",
                       "2015-12-08 00:00:00", 
                       "2016-01-01 00:00:00", 
                       "2016-01-02 00:00:00"),
                     tz = "GMT"),
  end = as.POSIXct(c("2016-01-05 00:00:00", 
                     "2015-12-19 03:30:00", 
                     "2016-01-01 05:00:00",
                     "2016-01-05 00:00:00"), 
                   tz = "GMT"),
  area = c(300, 200, 200, 450)
)
```
`r knitr::kable(deploy, caption = "deploy version 2")`

Each row represents a period of **continuous functionality** for a camera. If a camera ever went offline then came back on, it should have multiple rows. For cameras with multiple entries, the time periods should never overlap. Whenever the camera's area changes, it should be entered as a new row. area = 0 is equivalent to the camera was not functioning. 

This version of deploy must have, at a minimum, these four columns:

- cam: The unique ID of the sampling site (same class as df$cam)

- start: the first time this camera with this area was active (class POSIXct, same timezone as df$datetime)

- end: the last time this camera with this area was active (class POSIXct, same timezone as df$datetime)

- area: the camera's visible area, in any units (class numeric)


## Time-to-Event
This is in progress. No code yet.

## Space-to-Event
### Specify occasions

The first step after defining df and deploy is to specify the sampling occasions. This can be done manually or with the function build_occ(). The sampling occasions should be in a data.frame or tibble with the following structure:
```{r}
study_dates <- as.POSIXct(c("2016-01-01 00:00:00", "2016-01-04 23:59:59"), tz = "GMT")
occ <- build_occ(samp_freq = 3600, # seconds between each sampling occasion
                 samp_length = 10, # duration of each sampling occasion (seconds)
                 study_start = study_dates[1],
                 study_end = study_dates[2])
```
`r knitr::kable(head(occ), caption = "example structure of occ")`


### Build your encounter history

With one simple step, you can build your space-to-event encounter history. 
Be sure to specify your study_area size in the same units as your camera visible areas 
```{r}
ste_eh <- build_ste_eh(df, deploy, occ)
ste_eh
```

### Estimate abundance
To estimate abundance from STE, use the function ste_estN_fn
```{r}
ste_estN_fn(ste_eh, study_area = 1e6)
```

This will always produce a warning about Nelder-Mead. In future versions of this package, you may be able to change your optimization method.


## Instantaneous Sampling.
### Specify occasions

As with STE, the first step is to specify the sampling occasions. This can be done manually or with the function build_occ(). The sampling occasions should be in a data.frame or tibble with the following structure:
```{r}
study_dates <- as.POSIXct(c("2016-01-01 00:00:00", "2016-01-04 23:59:59"), tz = "GMT")
occ <- build_occ(samp_freq = 3600,
                 samp_length = 10,
                 study_start = study_dates[1],
                 study_end = study_dates[2])
```
`r knitr::kable(head(occ), caption = "example structure of occ")`

### Build your encounter history
```{r}
ise_eh <- ise_build_eh(df, deploy, occ, assume0 = T) 
head(ise_eh)
```

### Estimate abundance 
```{r}
ise_estN <- ise_estN_fn(
  ise_eh,
  A = 1e6
)
ise_estN
```
