---
title: "Using spaceNtime"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{spaceNtime}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(dplyr)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The package spaceNtime was created to assist with analyses described in ["Three novel methods to estimate abundance of unmarked animals using remote cameras" by Moeller et al. (2018)](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1002/ecs2.2331). This package helps the user create an encounter history from a database of analyzed photos and run each of the models.

Currently, this package only includes STE, but the other methods are on their way. 

This is my first package and is very much in development, so it is guaranteed to have bugs. I would appreciate hearing from you if you encounter problems!

## Load the package
If you haven't already installed it, please see instructions at the bottom of my [Github repository](https://github.com/annam21/spaceNtime)
```{r, message = F}
library(spaceNtime)
```


## The data

The functions for all three methods require two pieces of information.

### Photo information

The first, called df, is a data.frame or tibble that contains all your pictures from the study period. This must contain, at a minimum, these three columns (names and classes must match exactly): 

- cam: The unique ID of the sampling site (any class)

- datetime: The date and time of the photo (class POSIXct)

- count: The count of your study species at each date and time (class numeric)

```{r}
df <- data.frame(
  cam = c(1,1,2,2,2),
  datetime = as.POSIXct(c("2016-01-02 12:00:00",
                          "2016-01-03 13:12:00",
                          "2016-01-02 12:00:00",
                          "2016-01-02 14:00:00",
                          "2016-01-03 16:53:42"),
                        tz = "GMT"),
  count = c(1, 0, 0, 1, 2)
)
```
`r knitr::kable(df, caption = "df")`

### Camera deployment information


The user must also provide a database called deploy. This is a dataframe or tibble that contains information about the active periods for each active camera and their areas. **Please note,** to get accurate estimates, this should include all cameras that were functioning, not just the ones that got pictures of your target species. 

#### Version 1. 

```{r}
deploy <- data.frame(
  cam = c(1, 2, 2, 2),
  start = as.POSIXct(c("2015-12-01 15:00:00",
                       "2015-12-08 00:00:00", 
                       "2016-01-01 00:00:00", 
                       "2016-01-02 00:00:00"),
                     tz = "GMT"),
  end = as.POSIXct(c("2016-01-05 00:00:00", 
                     "2015-12-19 03:30:00", 
                     "2016-01-01 05:00:00",
                     "2016-01-05 00:00:00"), 
                   tz = "GMT"),
  area = c(300, 200, 200, 450)
)
```
`r knitr::kable(deploy, caption = "deploy version 2")`

This version can be used for either motion-sensor or timelapse photos. Each row represents a period of **continuous functionality** for a camera. If a camera ever went offline then came back on, it should have multiple rows. For cameras with multiple entries, the time periods should never overlap. Whenever the camera's area changes, it should be entered as a new row. Setting area = 0 is equivalent to the camera not functioning. If a camera has any missing periods, it will be assumed that the camera was not working during that time.  

#### Version 2: Timelapse only
Users with timelapse data have the option of creating deploy directly from df. This is the suggested form for timelapse photos because any missing photos are interpreted as camera malfunctions rather than counts of 0 animals. Furthermore, this allows the most flexibility for users with occasion-by-occasion information about the camera's area and functionality collected collected while analyzing the photos. For example, a photo with the lens covered in snow could be coded as area = 0. 

```{r, echo = F}
df1 <- data.frame(
  cam = c(1,1,2,2,2),
  datetime = as.POSIXct(c("2016-01-02 12:00:00",
                      "2016-01-02 14:00:00",
                      "2016-01-02 12:00:00",
                      "2016-01-02 14:00:00",
                      "2016-01-03 16:00:00"),
                    tz = "GMT"),
  count = c(0, 1, 0, 0, 2),
  area = c(300, 300, 200, 0, 450)
)
deploy1 <- df1 %>%
  mutate(start = datetime, 
         end = start) %>%
  select(cam, start, end, area)
```
`r knitr::kable(deploy1, caption = "deploy occasion-by-occasion")`

This version of deploy should contain *every* functional photo (i.e., not obstructed or malfunctioning) taken during the study period. It will have the following columns, at a minimum:

- cam: The unique ID of the sampling site (same class as df$cam)

- start: The date and time of the photo (same class and tz as df$datetime)

- end: The date and time of the photo (same class and tz as df$datetime)

- area: the camera's visible area, in any units (class numeric)

The area should be 0 on any occasion that the camera was obstructed or malfunctioning. 

## Time-to-Event
This is in progress. No code yet.

## Space-to-Event
### Specify occasions

The first step after defining df and deploy is to specify the sampling occasions. This can be done manually or with the function build_occ(). The sampling occasions should be in a data.frame or tibble with the following structure:
```{r}
study_dates <- as.POSIXct(c("2016-01-01 00:00:00", "2016-01-04 23:59:59"), tz = "GMT")
occ <- build_occ(samp_freq = 3600, # seconds between the start of each sampling occasion
                 samp_length = 10, # duration of each sampling occasion (seconds)
                 study_start = study_dates[1],
                 study_end = study_dates[2])
```
`r knitr::kable(head(occ), caption = "example structure of occ")`

### Build your encounter history

With one simple step, you can build your space-to-event encounter history. 

```{r}
ste_eh <- ste_build_eh(df, deploy, occ)
ste_eh
```

### Estimate abundance

To estimate abundance from STE, use the function ste_estN_fn. 
Be sure to specify your study_area size in the same units as your camera visible areas. 

```{r}
ste_estN_fn(ste_eh, study_area = 1e6)
```

## Instantaneous Sampling
### Specify occasions

As with STE, the first step is to specify the sampling occasions. This can be done manually or with the function build_occ(). The sampling occasions should be in a data.frame or tibble with the following structure:
```{r}
study_dates <- as.POSIXct(c("2016-01-01 00:00:00", "2016-01-04 23:59:59"), tz = "GMT")
occ <- build_occ(samp_freq = 3600,
                 samp_length = 10,
                 study_start = study_dates[1],
                 study_end = study_dates[2])
```
`r knitr::kable(head(occ), caption = "example structure of occ")`

### Build your encounter history


**NOTE** The assmume_p1 argument is crucial for getting accurate results. To build the encounter history, this code will look for photos at each camera during each occasion. If there are no photo there, should the model assume that no animals were present or that the camera malfunctioned?

- assume_p1 = T: The camera did not take a photo because there were no animals present (For motion sensor photos; assuming p = 1) 

- assume_p1 = F: The camera did not take a photo because it stopped functioning (For timelapse photos; assuming it was supposed to take a photo at every occasion)

```{r}
ise_eh <- ise_build_eh(df, deploy, occ) 
head(ise_eh)
```

### Estimate abundance 
```{r}
ise_estN_fn(
  ise_eh,
  study_area = 1e6
)
```
